---
title: "MA589 - Computational Statistics Project 2"
author: "Megha Pandit"
date: "October 9, 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1.(a) Explain why horner works; start by understanding how a small polynomial, say c(3, -2, 1), is evaluated, and then provide a mathematical expression that summarizes how the computations are performed.
```{r}
horner <- function (coef)
function (x) {
s <- 0
for (i in seq(length(coef), 1, -1)){
s <- s * x + coef[i]
}
s
}
coef <- c(3,-2,1)
x <- 3
horner(coef)(x)
```

**_Evaluating the polynomial c(3,-2,1) at x = 3, the above function computes it iteratively for i = 3,2 and 1, as_** $s_2 = s_2x + coef_3$, $s_1 = s_1x + coef_2$ **_and_** $s_0 = s_0x + coef_1$. **_Mathematically, if_** $p(x)$ **_is a polynomial_** $p(x) = b_0 + b_1x + b_2x^2 + b_3x^3 + ... + b_nx^n$, **_then, we can evaluate_** $p(x)$ **_at_** $x = x_0$ **_as follows:_**
**_We can write the polynomial as_** 
$$p(x_0) = b_0 + x_0(b_1 + x_0(b_2 + x_0(....x_0(b_{n-1} + b_nx_0))))$$
**_We can define constants_** $s_n, s_{n-1},...$ **_such that_** 
$$s_n = s_nx_0 + b_n$$
**_Since we are starting with s = 0,_**  
$$s_n = 0 + b_n = b_n \\s_{n-1} = s_nx_0 + b_{n-1} \\s_{n-2} = s_{n-1}x_0 + b_{n-2} \\.\\.\\.\\s_0 = s_1x_0 + b_0$$
**_Iteratively substituting_** $s_i, i = n,(n-1),(n-2),...,0$,**_in the p(x) equation, we get_** $p(x_0) = s_0$.
**_In short, we can write the mathematical expression as: for_** $i = n, (n-1), (n-2),...,0, \\s_i = b_i$ **_and_** $s_{i-1} = s_ix + b_{i-1}$.

1.(b) Use horner to plot $p(x) = 2 - 4x - x^2/2 + x^3$ for $x \in [-3,3]$. (Hint: check the code in the next problem, or see curve in R.)
```{r}
c <- c(2,-4,-1/2,1)
curve(horner(c)(x), from = -3, to = 3)
```

1.(c) Implement Newton’s method to find the roots of p from the previous item. 
```{r}
#horner function for evaluating polynomial
horner <- function (coef)
function (x) {
s <- 0
for (i in seq(length(coef), 1, -1)){
s <- s * x + coef[i]
}
s
}

#horner function for evaluating the derivative of polynomial
dhorner <- function(coef)
function(x){
s <- 0
for (i in seq(length(coef), 2, -1)){
  s <- s*x + (i-1)*coef[i]
  }
  s
}

#Newton's method for finding the roots of the polynomial
newton <- function(x){
  p <- x
  m <- p - horner(coef)(p)/dhorner(coef)(p)
  while (abs(p-m) > 1e-12) {
    p <- m
    m <- p - horner(coef)(p)/dhorner(coef)(p)
  }
  print(p)
}
coef <- c(2,-4,-1/2,1)
newton(-1.5)
newton(0)
newton(1.5)
```
**_Starting from x = -1 gives an error._**


1.(d) Legendre polynomial
```{r, include=FALSE}
legendre <- function(n){
  M <- matrix(0, nrow = n+1, ncol = n+1)
  M1 <- matrix(0, nrow = n+1, ncol = n+1)
  #When n = 0, M[1,1]=1
  M[1,1] <- 1
  M[2,1:2] <- c(0,1)
    for (i in 2:n) {
      M1 <- append(M[i,], values = 0, after = 0)
      M[i+1,] <- ((2*i + 1)*M1[-(n+2)] - (i-1)*M[i-1,])/i
    }
  return(M[n,])
}
legendre(4)
```

2.(a)creating the tableau T
```{r}
#creating a vandermonde matrix of predictors
T <- matrix(0,6,6)
x <- c(-3.0, -2.5, -2.0, -1.5, -1.0, -0.5, 0.0, 0.5, 1.0, 1.5, 2.0, 2.5, 3.0)
X <- outer(x, 0:4, FUN = "^")

#creating the tableau T
y <- c(-17.0, -7.0, 0.5, 3.0, 5.0, 4.0, 2.5, -0.5, -2.0, -2.0, 0.5, 5.0, 12.0)
T[1:5,1:5] <- crossprod(X,X)
T[1:5,6] <- crossprod(X,y)
T[6,1:5] <- crossprod(y,X)
T[6,6] <- crossprod(y,y)
T

```

2.(b)
```{r}
SWEEP <- function(T, k){
  n <- nrow(T)
  D <- T[k,k]
  T[k,] <- T[k,]/D
  for (i in 1:n) {
    if (i != k){
      B <- T[i,k]
      T[i,] <- T[i,] - B*T[k,]
      T[i,k] <- (-1)*B/D
    }
  }
  T[k,k] <- 1/D
  return(T)
}

for (i in 1:5) {
  T <- SWEEP(T,i)
}
T
```

**_#Quick Check: SWEEP(SWEEP(T, k), k) returns the original tableau T._**
```{r}
#Quick Check:
for (i in 1:5) {
SWEEP(SWEEP(T, i), i)
}
T
```

2.(c)
```{r}
T <- matrix(0,6,6)
x <- c(-3.0, -2.5, -2.0, -1.5, -1.0, -0.5, 0.0, 0.5, 1.0, 1.5, 2.0, 2.5, 3.0)
X <- outer(x, 0:4, FUN = "^")

#creating the tableau T
y <- c(-17.0, -7.0, 0.5, 3.0, 5.0, 4.0, 2.5, -0.5, -2.0, -2.0, 0.5, 5.0, 12.0)
T[1:5,1:5] <- crossprod(X,X)
T[1:5,6] <- crossprod(X,y)
T[6,1:5] <- crossprod(y,X)
T[6,6] <- crossprod(y,y)
T
plot(x, y)
a <- seq(-3, 3, length.out = 100)
p <- ncol(T) - 1
for (k in 1:4) {
T <- SWEEP(T, k)
lines(a, horner(T[1:k, p + 1])(a), lty = k)
print(c(k, T[p + 1, p + 1]))
}
```
**_For each k, the regression line with degree k for_** $y = \sum\limits_{i = 0}^n\beta_ix^i$ **_is getting plotted. For k=1, it is a straight line. For k=2, it is a linear regression line. For k=3, it is a parabola and for k=4, it is a curve._**
**_The T[p+1,p+1] element is the RSS. Therefore, for each k, the RSS is being printed._**

3.(a)Inverse Schur Complement
```{r}
P <- matrix(c(4,5,3,1),2,2) 
Q <- matrix(c(3,5,4,7,2,1),2,3)
R <- matrix(c(1,8,6,4,3,2),3,2)
S <- matrix(c(1,3,4,2,5,6,1,9,5),3,3)

K <- rbind(cbind(P, Q), cbind(R,S))
solve(K)

#inverse Schur complement
S1 <- K[3:5,3:5] - K[3:5,1:2]%*%solve(K[1:2,1:2])%*%K[1:2,3:5]
solve(S1)

for (k in 1:2) {
  K <- SWEEP(K,k)
}
K

for (k in 3:5) {
 K <- SWEEP(K,k) 
}
K #The yy block of K is equal to the inverse Schur complement
```
